---
title: "Template Cox proportional hazards model"
output: html_notebook
author: "João Vítor Souza"
---

Packages required
```{r}
# Essential packages 
library(survival); library(survminer)
# Useful packages
library(gtsummary); library(ggplot2); library(dplyr)
```

## Sample data set

The lung dataset is available from the `survival` package in R. The data contain subjects with advanced lung cancer from the North Central Cancer Treatment Group (NCCTG). 

Variables of interest:

* time: Survival time in days;
* status: censoring status 1=censored, 2=dead;
* sex: Male=1 Female=2;

```{r}
data(lung)
head(lung)
```
## Check the distribution of time of survival

Here we will also check for the differences between patients that were censored or not
```{r, message=FALSE, echo=FALSE, warnig=FALSE}
lung %>% mutate(Status=factor(status, levels=c(1,2), labels=c("Censored", "Dead"))) %>%
  ggplot(aes(x = time, fill = Status)) + 
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") + 
  labs(x = "Days survived", y = "N. of patients")
```

## What if you need to calculate the time?

It is common for us to only have dates that events occurred. In this sense, we can work with the date/time format to get the difference of time between these events.

Examples:
* Date and time of death - Date and time of beginning of treatment;
* Date and time of discharge - Date and time of admission;

### Example 1: Dates

Here I will create a small example of dates of arrival (`date_arred`) and discharge (`dc_date`) from hospital.
```{r}
date_ex <- 
  tibble(date_arred = c("2007-02-20", "2004-08-13", "2010-10-27"), 
         dc_date = c("2007-04-15", "2004-08-18", "2010-10-28"))
date_ex
```

We can transform dates into `POSIXct` format with the as.POSIXct function.

Here we can specify the format (e.g. YYYY-mm-dd) of the date/time variables as well as the separators used. 

[List of available formats - bottom of the page](https://stat.ethz.ch/R-manual/R-devel/library/base/html/strptime.html)

```{r}
# Transforming variables to date/time format
arrival <- as.POSIXct(date_ex$date_arred, format="%Y-%m-%d")
discharge <- as.POSIXct(date_ex$dc_date, format="%Y-%m-%d")

# Calculating the difference of time between arrival and discharge
date_ex$delta_arr_dc <- as.numeric(difftime(discharge, arrival, units = "days"))

date_ex
```

### Example 2: Dates and times

Now, let's assume you have both date and time of arrival and discharge and want to consider time as well.

```{r}
date_ex2 <- 
  tibble(date_arred = c("2007-02-20", "2004-08-13", "2010-10-27"),
         time_arred = c("20:20", "18:00", "12:15"),
         dc_date = c("2007-04-15", "2004-08-18", "2010-10-28"),
         dc_time = c("19:11", "10:10", "09:00"))
date_ex2
```

Now we need to:
1. Paste date and time together;
2. Transform it to POSIXct considering the new format;
3. Calculate the differences;

```{r}
# Here I paste the date and time of arrival into a single thing
arrival_date_time <- paste(date_ex2$date_arred, date_ex2$time_arred, sep="-")
arrival_date_time
# Now I transform it to date/time, considering the format needed
arrival_date_time <- as.POSIXct(arrival_date_time, 
                                format="%Y-%m-%d-%H:%M")
# Now we do the same for discharge
discharge_date_time <- paste(date_ex2$dc_date, date_ex2$dc_time, sep="-")
discharge_date_time

discharge_date_time <- as.POSIXct(discharge_date_time, 
                                format="%Y-%m-%d-%H:%M")

# Calculating the difference of time between arrival and discharge
date_ex2$delta_arr_dc <- as.numeric(difftime(discharge_date_time,
                                            arrival_date_time, 
                                            units = "days"))
# Now we have the amount of days between arrival and discharge, considering the time as well
date_ex2
```

## Kaplan-Meier method

```{r}
f1 <- survfit(Surv(time, status) ~ 1, data = lung)
f1
```

```{r}
surv.f1 <- ggsurvplot(
  fit = f1, 
  xlab = "Days", 
  ylab = "Overall survival probability")

plot.new() 
print(surv.f1,newpage = FALSE)
```

## Comparing survival times between groups


```{r}
lung$sex1 <- factor(lung$sex, levels=c(1,2), labels=c("Male", "Female"))


fit <- survfit(Surv(time, status) ~ sex1 , data = lung)

surv.sex <- ggsurvplot(fit=fit, data = lung, 
                       linetype="strata", surv.median.line = "hv",
                       conf.int = TRUE, pval=TRUE)

plot.new() 
print(surv.sex,newpage = FALSE)
```
### Cox proportional hazards model

Testing the effect of a single variable
```{r}
res.cox <- coxph(Surv(time, status) ~ sex1 , data = lung)
res.cox %>% tbl_regression(exp = TRUE) %>% bold_p()
```


### Testing multiple variables at once
```{r}
# Changing variable to factor
lung$ph.ecog1 <- factor(lung$ph.ecog)
# Include the variables you want to test with the time to event
covariates <-  c("sex", "age", "ph.karno", "ph.ecog1")
# Use this list to include nice labels for the final table
var_names <- list(sex~"Sex",
                  age~"Age",
                  ph.karno~"PH Karno",
                  ph.ecog1~"PH Ecog")

tbl_univ <-
  tbl_uvregression(
    lung[,c(covariates, "time", "status")],
    label=var_names,
    method = coxph,
    y = Surv(time, status),
    exponentiate = TRUE) %>% 
  bold_p() %>% bold_labels() 
tbl_univ
```
### Multi-variate Cox proportonional hazards model
```{r}
var_names <- list(sex~"Sex",
                  age~"Age",
                  ph.karno~"PH Karno",
                  ph.ecog1~"PH Ecog")

tbl_multiv <- coxph(Surv(time, status) ~ sex + age + ph.karno + ph.ecog1, data = lung) %>% 
  tbl_regression(label=var_names, exponentiate=TRUE) %>% bold_p() %>% bold_labels
tbl_multiv

```
Combining tables
```{r}
tbl_merge(tbls=list(tbl_univ, tbl_multiv),
          tab_spanner=c("Univariate", "Multivariate"))
```



