---
title: "Template Cox proportional hazards model"
output: html_notebook
author: "João Vítor Souza"
---

Packages required
```{r, message=FALSE, warning=FALSE}
# Essential packages 
library(survival); library(survminer)
# Useful packages
library(gtsummary); library(ggplot2); library(dplyr)
```

## Sample data set

The lung dataset is available from the `survival` package in R. The data contain subjects with advanced lung cancer from the North Central Cancer Treatment Group (NCCTG). 

Variables of interest:

* time: Survival time in days;
* status: censoring status 1=censored, 2=dead;
* sex: Male=1 Female=2;

```{r}
head(lung)
```


## Check the distribution of time of survival

Here we will also check for the differences between patients that were censored or not
```{r, message=FALSE, echo=FALSE, warnig=FALSE}
lung %>% mutate(Status=factor(status, levels=c(1,2), labels=c("Censored", "Dead"))) %>%
  ggplot(aes(x = time, fill = Status)) + 
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") + 
  labs(x = "Days survived", y = "N. of patients")
```

## Censoring

## What if you need to calculate the time?

It is common for us to only have dates that events occurred. In this sense, we can work with the date/time format to get the difference of time between these events.

Examples:
* Date and time of death - Date and time of beginning of treatment;
* Date and time of discharge - Date and time of admission;

### Example 1: Dates

Here I will create a small example of dates of arrival (`date_arred`) and discharge (`dc_date`) from hospital.
```{r}
date_ex <- 
  tibble(date_arred = c("2007-02-20", "2004-08-13", "2010-10-27"), 
         dc_date = c("2007-04-15", "2004-08-18", "2010-10-28"))
date_ex
```

We can transform dates into `POSIXct` format with the as.POSIXct function.

Here we can specify the format (e.g. YYYY-mm-dd) of the date/time variables as well as the separators used. 

[List of available formats - bottom of the page](https://stat.ethz.ch/R-manual/R-devel/library/base/html/strptime.html)

```{r}
# Transforming variables to date/time format
arrival <- as.POSIXct(date_ex$date_arred, format="%Y-%m-%d")
discharge <- as.POSIXct(date_ex$dc_date, format="%Y-%m-%d")

# Calculating the difference of time between arrival and discharge
date_ex$delta_arr_dc <- as.numeric(difftime(discharge, arrival, units = "days"))

date_ex
```

### Example 2: Dates and times

Now, let's assume you have both date and time of arrival and discharge and want to consider time as well.

```{r}
date_ex2 <- 
  tibble(date_arred = c("2007-02-20", "2004-08-13", "2010-10-27"),
         time_arred = c("20:20", "18:00", "12:15"),
         dc_date = c("2007-04-15", "2004-08-18", "2010-10-28"),
         dc_time = c("19:11", "10:10", "09:00"))
date_ex2
```

Now we need to:
1. Paste date and time together;
2. Transform it to POSIXct considering the new format;
3. Calculate the differences;

```{r}
# Here I paste the date and time of arrival into a single thing
arrival_date_time <- paste(date_ex2$date_arred, date_ex2$time_arred, sep="-")
arrival_date_time
# Now I transform it to date/time, considering the format needed
arrival_date_time <- as.POSIXct(arrival_date_time, 
                                format="%Y-%m-%d-%H:%M")
# Now we do the same for discharge
discharge_date_time <- paste(date_ex2$dc_date, date_ex2$dc_time, sep="-")
discharge_date_time

discharge_date_time <- as.POSIXct(discharge_date_time, 
                                format="%Y-%m-%d-%H:%M")

# Calculating the difference of time between arrival and discharge
date_ex2$delta_arr_dc <- as.numeric(difftime(discharge_date_time,
                                            arrival_date_time, 
                                            units = "days"))
# Now we have the amount of days between arrival and discharge, considering the time as well
date_ex2
```

## Kaplan-Meier method

```{r}
f1 <- survfit(Surv(time, status) ~ 1, data = lung)
f1
```

```{r}
surv.f1 <- ggsurvplot(
  fit = f1, 
  xlab = "Days", 
  ylab = "Overall survival probability")

plot.new() 
print(surv.f1,newpage = FALSE)
```

## Comparing survival times between groups


```{r}
lung$sex1 <- factor(lung$sex, levels=c(1,2), labels=c("Male", "Female"))

# specify event 1=event at time, 2=left censored, 3=interval censored
fit <- survfit(Surv(time, status) ~ sex1 , data = lung)

surv.sex <- ggsurvplot(fit=fit, data = lung, 
                       linetype="strata", surv.median.line = "hv",
                       conf.int = TRUE, pval=TRUE)

plot.new() 
print(surv.sex,newpage = FALSE)
```
### Cox proportional hazards model

http://www.sthda.com/english/wiki/cox-proportional-hazards-model 

```{r}
res.cox <- coxph(Surv(time, status) ~ sex1 , data = lung)
res.cox
```

Interpret raw results
```{r}

```

Testing the effect of a single variable
```{r}

res.cox %>% tbl_regression(exp = TRUE) %>% bold_p()
```


### Testing multiple variables at once
```{r}
# Changing variable to factor
lung$ph.ecog1 <- factor(lung$ph.ecog)
# Include the variables you want to test with the time to event
covariates <-  c( "age","sex1", "wt.loss", "ph.ecog1")
# Use this list to include nice labels for the final table
var_names <- list(age~"Age",
                  sex1~"Sex",
                  wt.loss~"Weight Loss",
                  ph.ecog1~"PH Ecog")

tbl_univ <-
  tbl_uvregression(
    lung[,c(covariates, "time", "status")],
    label=var_names,
    method = coxph,
    y = Surv(time, status),
    exponentiate = TRUE) %>% 
  bold_p() %>% bold_labels() 
tbl_univ
```
### Multi-variate Cox proportonional hazards model
```{r}
res.cox.multiv <- coxph(Surv(time, status) ~  age + sex1 + wt.loss + 
                        ph.ecog1, data = lung) 
res.cox.multiv
```

Generating a nice table of results
```{r}
var_names <- list(age~"Age",
                  sex1~"Sex",
                  wt.loss~"Weight Loss",
                  ph.ecog1~"PH Ecog")
tbl_multiv <- res.cox.multiv %>% 
  tbl_regression(label=var_names, exponentiate=TRUE) %>% bold_p() %>% bold_labels
tbl_multiv

```


Combining tables
```{r}
tbl_merge(tbls=list(tbl_univ, tbl_multiv),
          tab_spanner=c("Univariate", "Multivariate"))
```

## Checking model assumptions

After we have built our model it is necessary to check wether it is biased or violates assumptions.

The tests we need to perform are:
* Proportional hazards assumption;
* Examining influential observations (or outliers).
* Detecting non linearity in relationship between the log hazard and the covariates.

In order to check for these, *Residuals* method are used. Common residuals for cox include:

* Schoenfeld residuals to check the proportional hazards assumption;
* Martingale residual to assess non linearity;
* Deviance residual (symmetric transformation of the Martinguale residuals), to examine influential observations;

[Complete guide - Sthda](http://www.sthda.com/english/wiki/cox-model-assumptions)

```{r}
res.cox.multiv
```
### Proportional hazard assumption

```{r}
test.ph <- cox.zph(res.cox.multiv)
test.ph
```

```{r}
plot.ph <- ggcoxzph(test.ph) 

plot.new()
print(plot.ph,newpage = FALSE)
```

Violations of proportional hazards assumption can be resolved by:

* Adding covariate * time interaction
* Stratification

```{r}
library(Greg)
```

```{r}
lung.update <- lung %>% 
  mutate(status= factor(status,
                        levels=1:2,
                        labels=c("Censored","Died")),
         time=time/365.25,
         sex=factor(sex,
                    levels=1:2,
                    labels=c("Male","Female")),
         ph.ecog=factor(ph.ecog)
         )
```

```{r}
regular.model <- coxph(Surv(time, status=="Died") ~ sex + age + wt.loss + ph.ecog, data = lung.update)

summary(regular.model)
```

```{r}

split_lung <- lung.update %>% timeSplitter(by=.5,
                                    event_var ="status",
                                    time_var="time",
                                    time_related_vars = "age")

interval.model <-
  update(regular.model, 
         Surv(Start_time, Stop_time, status == "Died") ~ .,
         data = split_lung)

summary(interval.model)
```

```{r}
library(htmlTable)
cbind(Regular = coef(regular.model),
      Interval = coef(interval.model),
      Difference = coef(regular.model) - coef(interval.model)) %>% 
  txtRound(digits = 5) %>% 
  knitr::kable(align = "r")
```

```{r}
cox.zph(regular.model) 
```
```{r}
time.int.model <- update(interval.model,
                         .~.+sex:Start_time)
summary(time.int.model )
```

```{r}
split_lung %<>% 
  mutate(sex_start = sex * Start_time) 
anova(time.int.model,
      update(time.int.model, .~.+I(sex_start^2)))

split_lung
```





### Examining influential observations
```{r, message=FALSE, warning=FALSE}
ggcoxdiagnostics(res.cox.multiv, type='dfbeta', linear.predictions = TRUE)
```



```{r}
ggcoxdiagnostics(res.cox.multiv, type = "deviance",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```


### Testing non-linearity

Necessary only for continuous covariates.

Martingale residuals may present any value in the range (-INF, +1):

* Residuals near 1 represents individuals that “died too soon”,
* and large negative values correspond to individuals that “lived too long”.

```{r, fig.height=5}
age.lin_plot <- ggcoxfunctional(Surv(time, status) ~ age + log(age) + sqrt(age), data = lung)
plot.new() 
print(age.lin_plot,newpage = FALSE)
```



