---
title: "Template Interrupted time series"
output: html_notebook
author: "João Vítor Souza"
---

Loading packages and sample data set.
```{r}
library(astsa); library(forecast); library(dplyr); library(zoo); library(comprehenr);
library(ggplot2); library(scales); library(gtsummary); library(foreign);library(tsModel); library("lmtest") ; library("Epi");library("splines") ; library("vcd")

data <- read.csv("sicily.csv")
```

## Sample data set


The **Sicily** dataset consistis of Acute Coronary Events (ACEs) registerd in Sicily befor and after the **ban of smoking in public places**. It is available as a suplementary material from the publication [Interrupted time series regression for the evaluation of public health interventions: a tutorial](https://academic.oup.com/ije/article/46/1/348/2622842#112547752).

### Variables of interest

* year: year of the event;
* month: coded in ordinal number going from 1 (Jan) to 12 (Dec);
* aces: sum of acute coronary events registered in that month;
* time: numeric variable continuously counting how many periods have passed since the beginning of the study;
* smokban: dummy variable indicating whether the period is before (0) or after (1) the ban of smoking in public places;
* pop: population of Sicily;
* stdpop: standardized population of Sicily.

```{r}
head(data)
```

## Explore ACE events

```{r}
### The months variable is coded as 1, 2 ... and I need it to be 01, 02..
numbers <- c()
for (i in as.character(data$month)){
  if (nchar(i) ==1){
    numbers <- c(numbers, paste("0",i, sep=""))
  }
  else {
    numbers <- c(numbers, i)
  }
}
# Creating a date column

data$date <- as.POSIXct(paste(data$year, numbers, "01",sep='-'), format="%Y-%m-%d")
data$smokban <- factor(data$smokban, levels=c(0,1),
                       labels=c("Before","After"))
```

Plotting the series of cases
```{r}
data %>% 
  ggplot(aes(x=date,y=aces, color=smokban)) + 
  geom_point() + scale_fill_discrete(name="") + ylim(0,1200) + labs(x="Date", y="Monthly ACE counts")
```

## Summarizing events

In the table below we do a simple comparison of median ACE events before and after the Ban of Smoking in public places. In addition to that, we create a variable that accounts for population growth (rate of ACEs/100,000). We can see how the difference is lost when we consider the growth of the population
```{r, warning=FALSE, message=FALSE}
data$rate <- with(data, aces/stdpop*10^5)

var_names <- list(rate~"Rate ACE",
                  aces~"Raw ACE")
stat <-  all_continuous() ~ c("{median} ({p25}, {p75})", 
                              "{min}, {max}", "{mean}, {sd}")
types <- list(all_continuous() ~ "continuous2")

data %>%
  dplyr::select(aces,rate,smokban) %>% 
  tbl_summary(by=smokban, label=var_names,statistic=stat, type=types,) %>% 
  add_p() %>% bold_labels() %>% bold_p() %>%
  as_gt() %>% gt::tab_header(gt::md("**ACE events before and after the smoke ban**"))
```


## Segmented regression

Simplest and easiest way to model an ITS. Briefly, segmented (or *piecewise*) regression fits two regressions: **before** and **after** the intervention. 

Notes:
* Here we are proposing a level change. This is why the smokban variable is used as a dummy that changes to 1 after the ban. Other assumptions should take different modelling;
* Here we use an offset as a log of the stdpop because the poisson distribution is not suited for a y variable that is a rate;
* The poisson distribution is ideal for TS as we usually work with counts. The Poisson distribution has a positive and integer basis as you could not have something like -2 or 3.5 cases registered in a period.

```{r}
model1 <- glm(aces ~ offset(log(stdpop)) + smokban + time, family=poisson, data)
summary(model1)
```
With the raw summary of the model we see that our estimate shows that ban of smoking caused a reduction of ~11% in the ACE events (p<0.001).

Making a pretty table
```{r}
model1 %>%
  tbl_regression(label=list(smokban~"Smoke ban", time~"Time")) %>%
  bold_p() %>%
  bold_labels() %>% as_gt() %>% gt::tab_header("Segmented regression model")
```

```{r}
datanew <- data.frame(stdpop=mean(data$stdpop),
                      smokban=factor(rep(c(0,1),c(36,24)),
                                     levels=c(0,1),
                                     labels=c("Before", "After")),
                      time= 1:60,
                      month=rep(01:12,5),
                      year=c(rep(2002, 12),
                             rep(2003, 12),
                             rep(2004, 12),
                             rep(2005, 12),
                             rep(2006, 12)))
pred1 <- predict(model1, type="response", datanew)

getDates <- function(month, year){
  numbers <- c()
  for (i in as.character(month)){
    if (nchar(i) ==1){
      numbers <- c(numbers, paste("0",i, sep=""))
    }
    else {
      numbers <- c(numbers, i)
    }
  }
  date <- as.POSIXct(paste(year, month, "01",sep='-'), format="%Y-%m-%d")
  return(date)
  }

datanew$date <- getDates(datanew$month, datanew$year)

datanew$aces <- pred1


# Counter-factual
data.cont <- data.frame(stdpop=mean(data$stdpop),
                        smokban=factor(rep(0,c(60)),
                                     levels=c(0,1),
                                     labels=c("Before", "After")),
                        time= 1:60,
                        month=rep(01:12,5),
                        year=c(rep(2002, 12),
                             rep(2003, 12),
                             rep(2004, 12),
                             rep(2005, 12),
                             rep(2006, 12)))
data.cont$date <- getDates(data.cont$month, data.cont$year)
data.cont$aces <- predict(model1, type="response", data.cont)
```

```{r}
data %>% 
  ggplot(aes(x=date,y=aces, color=smokban)) +
  geom_point() + 
  geom_line(data=datanew) + 
  geom_line(data=data.cont, linetype = "dashed") +
  scale_fill_discrete(name="") + ylim(0,1200) + labs(x="Date", y="Monthly ACE counts")
```




