---
title: "Handling missingness in data"

author: "@souzajvp"
institute: "GEMINI data science meeting"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: css/custom-en.css
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: false
---

class: inverse, middle, center

# Imputation with `mice`

.footnote[
See [Codes](https://github.com/gemini-duke/toolbox/tree/main/missingness_analysis), [Video-TBA]() and [SOP-TBD]() and [miceVignettes](https://www.gerkovink.com/miceVignettes/) for materials on this section
]

---

## The `mice` package  

Designed for flexible imputation of incomplete data. 

--

### Working with mice

1. Load packages
```{r, warning=FALSE}
require(mice); require(lattice); require(gtsummary)
set.seed(123)
```

Sample data set
```{r, echo=FALSE}
knitr::kable(head(nhanes), format = 'html')
```

---

### Inspect the missing data pattern

Missing data matrix
```{r, fig.align="center", fig.height=6}
md_pattern <- md.pattern(nhanes)
```

---

class: inverse, middle, center

## Ad Hoc imputation methods

---

### Imputing with the mean

```{r, message=FALSE, warning=FALSE}
imp <- mice(nhanes, method='mean', m=1, maxit=1, print=FALSE)
```

```{r, echo=FALSE}
knitr::kable(head(complete(imp)), format = 'html')
```

---

### Imputing with the mean

```{r, echo=FALSE}
og.model <- lm(age~bmi, data=nhanes) %>% tbl_regression()
imp.model <- lm(age~bmi, data=complete(imp)) %>% tbl_regression()

merged.model <- tbl_merge(list(og.model, imp.model), tab_spanner = c("Og. data", "Mean-imputed")) 
merged.model %>% as_gt() %>% gt::tab_header(gt::md("**Age~BMI model**"))
```

No actual difference, as bmi follows a *normal-like* distribution

---

### Regression imputation

```{r,echo=TRUE}
imp <- mice(nhanes, method="norm.predict", m=1, maxit=1, print=FALSE)
```

```{r, echo=FALSE}
knitr::kable(head(complete(imp)), format = 'html')
```

---

### Regression imputation.


```{r, echo=FALSE}
imp.model <- lm(age~bmi, data=complete(imp)) %>% tbl_regression() %>% bold_p()

merged.model <- tbl_merge(list(og.model, imp.model), tab_spanner = c("Og. data", "Regression-imputed")) 
merged.model  %>%  as_gt() %>% gt::tab_header(gt::md("**Age~BMI model**"))
```

---

### Multiple imputation

The standard mice algorithm runs for 5 iterations and generates 5 imputations each.
```{r}
imp <- mice(nhanes, print=FALSE)
```

Checking the attributes of the `imp` object, we can access many different aspects of the process.
```{r}
attributes(imp)
```



---

### Multiple imputation

The original data is store as `imp$data` and the imputations with `imp$imp`;

We can return the imputed data set either in long or broad formats:
```{r}
c.long <- complete(imp, "long")
knitr::kable(head(c.long), format = 'html')
```

---

### Predictor matrix

`mice` allows you to explicitily determine which variables are going to be used for imputing others;

--

This is done through the predictor matrix
```{r}
pred <- imp$pred
knitr::kable(pred, format = 'html')
```

---

Let's say you want to remove `hyp` from the set of predictors, but want it to be predicted by other variables
```{r}
pred[, "hyp"] <- 0
pred
```

Now you can use your new matrix in the imputation process
```{r}
imp <- mice(nhanes, pred=pred, prinf=FALSE)
```

